Cartograms with Dynamic Features

### 2022-07-16 paper/section/appendix.tex
draft: appendix on QGIS pre-processing

@@

    \begin{algorithm}[h!]
        \caption{Procedure to derive a point based an edge and a distance.}\label{alg:derive corridor point}
        \textbf{Input:} \\
        $ \Edge \gets $ the edge used to derive the new point \\
        $ \Distance \gets $ the distance between $ \PointP $ and $ \EdgeStart $ \\

@@

\begin{algorithm}[h!]
    \caption{Procedure to derive an edge, $ \EdgeParallel $, that is parallel to $ \Edge $ with a distance of $ \Distance $.}\label{alg:derive corridor edge}

>>>>

\section{Pre-processing Shapefiles}

Shapefiles from different sources are likely to be incompatible. In our case, the NHS CCG shapefile is incompatible with the river shapefiles. The major reason for the incompatibility is the coordinate reference system (CRS). The CRS of the CCG shapefile is EPSG:27700 (OSGB36 - British National Grid). The CRS of the river shapefiles is EPSG:4326 (WGS84 - World Geodetic System). Here we provide some pre-processing steps using QGIS (version: 3.26.0-Buenos Aires) \cite{qgisWelcome} to handle the incompatibility issue and reduce shapefile size to improve performance.

\subsection{Import Shapefiles into QGIS}
We first load all three river shapefiles into QGIS \autoref{fig:import_rivers}, followed by the CCG shapefile \autoref{fig:import_ccgs}.

{
    \begin{figure}[tbh!]
        \centering
        \includegraphics[width=\columnwidth]{figure/qgis/import_rivers.png}
        \caption{QGIS interface, with River Trent, River Great Ouse, and River Thames (from top to bottom) imported.}
        \label{fig:import_rivers}
    \end{figure}

    \begin{figure}[tbh!]
        \centering
        \includegraphics[width=\columnwidth]{figure/qgis/import_ccgs.png}
        \caption{QGIS interface, with all NHS CCGs imported.}
        \label{fig:import_ccgs}
    \end{figure}

    \begin{figure}[tbh!]
        \centering
        \includegraphics[width=\columnwidth]{figure/qgis/unify_crs.png}
        \caption{QGIS interface, showing the unified CRS (OSGB36) for both layers.}
        \label{fig:unify_crs}
    \end{figure}
}

\subsection{Export Shapefiles in GeoJSON and Unify the Coordinate Reference System (CRS)}

We then use QGIS to unify the CRS, and export both layers in the GeoJSON format. See \autoref{fig:export_rivers} and \autoref{fig:export_ccgs}. The unified layer is shown in \autoref{fig:unify_crs}.

{
    \begin{figure}[tbh!]
        \centering
        \includegraphics[width=\columnwidth]{figure/qgis/export_rivers.png}
        \caption{QGIS interface, exporting all rivers using the OSGB36 CRS in GeoJSON.}
        \label{fig:export_rivers}
    \end{figure}

    \begin{figure}[tbh!]
        \centering
        \includegraphics[width=\columnwidth]{figure/qgis/export_ccgs.png}
        \caption{QGIS interface, exporting all NHS CCGs using the OSGB36 CRS in GeoJSON.}
        \label{fig:export_ccgs}
    \end{figure}

}

\subsection{Merge Shapefiles and Reduce File Size}

We then merge two layers into one layer, and export it in the TopoJSON format using Mapshaper \cite{blochMapshaper}. Mapshaper also supports the simplification of GeoJSON shapefiles. See \autoref{fig:export_topojson}.

{
    \begin{figure}[tbh!]
        \centering
        \includegraphics[width=\columnwidth]{figure/qgis/export_topojson.png}
        \caption{Mapshaper interface, merging all rivers with NHS CCGs into one layer, and export the merged layer in TopoJSON.}
        \label{fig:export_topojson}
    \end{figure}
}

\subsection{Pre-processing Result}

\autoref{tbale:pre-processing_result} shows the pre-processing result. The reduction in file size is significant and greatly reduces the initialization time of our implementation.

{
\renewcommand{\arraystretch}{1.5}
\begin{table}[!tb]
	\centering
	\resizebox{\columnwidth}{!}{
		\begin{tabulary}{\columnwidth}{|*{4}{l|}}
			\hhline{~|*{3}{-}}
			\multicolumn{1}{c|}{\textbf{Shapefile}} &
			\cellcolor{Mycolor2}\textbf{Original} &
			\cellcolor{Mycolor2}\textbf{GeoJSON} &
			\cellcolor{Mycolor2}\textbf{TopoJSON} \\
			\hline
			Rivers & 2.0 MB (GeoJSON) & 1.4 MB & -  \\
			\hline
			NHS CCGs & 46.6 MB (.shp, Esri vector shapefile) & 140.2 MB & -  \\
			\hline
			Merged & - & - & 16.3 MB \\
			\hline
		\end{tabulary}
	}
	\caption{The file size is reduced by 88.5\% from the original size.}
	\label{table:pre-processing_result}
\end{table}
}

\clearpage

@@

    \begin{algorithm}[tbh!]
        \caption{Procedure to derive a point based an edge and a distance.}\label{alg:derive corridor point}
        \textbf{Input:} \\
        $ \Edge \gets $ the edge used to derive the new point \\
        $ \Distance \gets $ the distance between $ \PointP $ and $ \EdgeStart $ \\

@@

\begin{algorithm}[tbh!]
    \caption{Procedure to derive an edge, $ \EdgeParallel $, that is parallel to $ \Edge $ with a distance of $ \Distance $.}\label{alg:derive corridor edge}

### 2022-07-16 paper/section/data.tex
draft: appendix on QGIS pre-processing

Geospatial boundaries, or shapefiles, were obtained from the sources described below.
{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=0.6\columnwidth]{figure/ccg.png}
    \caption{A map of 135 CCGs in England as of 2020, obtained from the Open Geography portalx \cite{opengeographyportalxOpen} with EPSG:27700 (OSGB36 - British National Grid) as the Coordinate Reference System (CRS).}
    \label{fig:ccg}
\end{figure}
}

\bobgraph{Clinical Commissioning Groups Shapefile:} Clinical Commissioning Groups (CCGs) are the primary administrative and geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to the NHS re-organization. The most up-to-date shapefile is available from the Open Geography portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, due to the absence of public EHR data published based on the latest CCG re-organization that took place in 2021.

>>>>

Geospatial boundaries, or shapefiles, were obtained from the sources described below.
{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=0.6\columnwidth]{figure/ccg.png}
    \caption{A map of 135 CCGs in England as of 2020, obtained from the Open Geography Portalx \cite{opengeographyportalxOpen} with EPSG:4326 (WGS84 - World Geodetic System) as the Coordinate Reference System (CRS).}
    \label{fig:ccg}
\end{figure}
}

\bobgraph{Clinical Commissioning Groups Shapefile:} Clinical Commissioning Groups (CCGs) are the primary administrative and geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to the NHS re-organization. The most up-to-date shapefile is available from the Open Geography Portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, due to the absence of public EHR data published based on the latest CCG re-organizations took place in 2021 and 2022.

### 2022-07-16 paper/section/introduction.tex
draft: appendix on QGIS pre-processing

Among the four types of cartograms categorized by \citea{nusrat2016State} (contiguous, non-contiguous, rectangular, and Dorling), a trade-off between types of accuracy is made (See \tableref{table:accuracy}). For this project we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. We implement a new cartographic layout algorithm that includes these topological features into the layout of the nodes representing geographical regions. To minimize geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm to a real world case-study using an Electronic Health Records (EHR) dataset to evaluate of our algorithm. We also present a user study demonstrating its effectiveness.

>>>>

Among the four types of cartograms categorized by \citea{nusrat2016State} (contiguous, non-contiguous, rectangular, and Dorling), a trade-off between types of accuracy is made (See \autoref{table:accuracy}). For this project we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. We implement a new cartographic layout algorithm that includes these topological features into the layout of the nodes representing geographical regions. To minimize geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm to a real world case-study using an Electronic Health Records (EHR) dataset to evaluate of our algorithm. We also present a user study demonstrating its effectiveness.

### 2022-07-16 paper/section/main.tex
draft: appendix on QGIS pre-processing

 {
  \begin{figure*}[tb!]
      \centering
      \includegraphics[width=\textwidth,height=0.95\textheight,keepaspectratio]{figure/flowchart.png}
      \caption{A flowchart illustration of our layout algorithm. See \figref{fig:flowchart-stalemate} for the flow of processing stalemate. See also \algoref{alg:UpdateLayout} for more detail. For illustration purposes, we show the rendered views alongside the logical views representing the actual computation.}
      \label{fig:flowchart}
  \end{figure*}

@@

\algoref{alg:UpdateLayout} and \figref{fig:flowchart} provide an overview of the process.

\bobgraph{Initialization:} We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a square node, $ \node $, with the initial size, $ \nodeSize $. We then load the river shapefiles and render the rivers. Since the vertices of the river in the shapefiles are not in sequential order, we first render the starting vertex, followed by the next closest vertex. This rendering approach enables us to adjust the river resolution as shown in \figref{fig:river resolution}. We further apply a smoothing logic by removing vertices that are too close to each other.

\bobgraph{Node layout:} The basic algorithm is as follows: we first apply the Fast Node Overlap Removal (FNOR) algorithm that solves the Variable Placement with Separation Constraints (VPSC) problem \cite{dwyer2006fast} in order to remove overlaps. We chose FNOR over other node overlap removal algorithms because FNOR is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. We gradually increase the node size by 1 unit at a time to ensure smooth transitions. An increase in $ \nodeSize $ can cause the nodes to overlap. During overlap removal, we compute node trajectories (See \algoref{alg:check river intersection}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. If a node oscillates across a river, we process this as a stalemate. This procedure ends when 1) no node overlap is present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached. The gradual size increase process provides stability to the layout and helps minimize geographical error.

@@

We use rivers as topological boundaries and prevent nodes from crossing them. The resolution of rivers can be adjusted by the user, as shown in \figref{fig:river resolution}, the number of vertices is reduced from 10,170 to 30. A pixel-level river resolution is not required as Demers cartograms do not need to represent geography at this level of detail. Instead a resolution that matches the node size is used. Enabling river simplification reduces the number of edge intersection tests that need to be performed. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \algoref{alg:check river intersection}. A bounding box intersection test between the edge defined by node translation and river edges can be performed to reduce the number of edge intersection tests required.

@@

As the FNOR always attempts to produce an optimal node layout where node distribution and translation are minimized, a node's translation path can repeatedly intersect a river due to congestion, creating a stalemate situation, as shown in \figref{fig:stalemate}. If a node is translated between two positions, $ \node $ and $ \nodeFNOR $, for $ \stalemateMax $ iterations (a user-adjustable parameter), we introduce a heuristic solution: constructing a corridor to alleviate the congestion. A corridor, $ \Corridor $, is a rectangle with a width of $ \CorridorWidth $ and a length of $ \CorridorLength $, formed by deriving two edges $ \EdgeParallelA $ and $ \EdgeParallelB $ such that $ \EdgeParallelA \parallel \EdgeParallelA \parallel \nodeLineNtNc $ (See \figref{fig:corridor}C and D). All nodes enclosed by $ \Corridor $ are then translated by $ \nodeVectorTC $ to alleviate the congestion (See \figref{fig:corridor}E).

@@

\begin{algorithm}[tb!]
    \caption{Procedure to derive a corridor to translate enclosed nodes. We use an SVG canvas, where the point of origin (0,0) is located at the top left corner, with the x-axis extending to the right and the y-axis extending downwards (See \figref{fig:stalemate}).}\label{alg:derive corridor}

>>>>

 {
  \begin{figure*}[tb!]
      \centering
      \includegraphics[width=\textwidth,height=0.95\textheight,keepaspectratio]{figure/flowchart.png}
      \caption{A flowchart illustration of our layout algorithm. See \autoref{fig:flowchart-stalemate} for the flow of processing stalemate. See also \autoref{alg:UpdateLayout} for more detail. For illustration purposes, we show the rendered views alongside the logical views representing the actual computation.}
      \label{fig:flowchart}
  \end{figure*}

@@

\autoref{alg:UpdateLayout} and \autoref{fig:flowchart} provide an overview of the process.

\bobgraph{Initialization:} We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a square node, $ \node $, with the initial size, $ \nodeSize $. We then load the river shapefiles and render the rivers. Since the vertices of the river in the shapefiles are not in sequential order, we first render the starting vertex, followed by the next closest vertex. This rendering approach enables us to adjust the river resolution as shown in \autoref{fig:river resolution}. We further apply a smoothing logic by removing vertices that are too close to each other.

\bobgraph{Node layout:} The basic algorithm is as follows: we first apply the Fast Node Overlap Removal (FNOR) algorithm that solves the Variable Placement with Separation Constraints (VPSC) problem \cite{dwyer2006fast} in order to remove overlaps. We chose FNOR over other node overlap removal algorithms because FNOR is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. We gradually increase the node size by 1 unit at a time to ensure smooth transitions. An increase in $ \nodeSize $ can cause the nodes to overlap. During overlap removal, we compute node trajectories (See \autoref{alg:check river intersection}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. If a node oscillates across a river, we process this as a stalemate. This procedure ends when 1) no node overlap is present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached. The gradual size increase process provides stability to the layout and helps minimize geographical error.

@@

We use rivers as topological boundaries and prevent nodes from crossing them. The resolution of rivers can be adjusted by the user, as shown in \autoref{fig:river resolution}, the number of vertices is reduced from 10,170 to 30. A pixel-level river resolution is not required as Demers cartograms do not need to represent geography at this level of detail. Instead a resolution that matches the node size is used. Enabling river simplification reduces the number of edge intersection tests that need to be performed. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \autoref{alg:check river intersection}. A bounding box intersection test between the edge defined by node translation and river edges can be performed to reduce the number of edge intersection tests required.

@@

As the FNOR always attempts to produce an optimal node layout where node distribution and translation are minimized, a node's translation path can repeatedly intersect a river due to congestion, creating a stalemate situation, as shown in \autoref{fig:stalemate}. If a node is translated between two positions, $ \node $ and $ \nodeFNOR $, for $ \stalemateMax $ iterations (a user-adjustable parameter), we introduce a heuristic solution: constructing a corridor to alleviate the congestion. A corridor, $ \Corridor $, is a rectangle with a width of $ \CorridorWidth $ and a length of $ \CorridorLength $, formed by deriving two edges $ \EdgeParallelA $ and $ \EdgeParallelB $ such that $ \EdgeParallelA \parallel \EdgeParallelA \parallel \nodeLineNtNc $ (See \autoref{fig:corridor}C and D). All nodes enclosed by $ \Corridor $ are then translated by $ \nodeVectorTC $ to alleviate the congestion (See \autoref{fig:corridor}E).

@@

\begin{algorithm}[tb!]
    \caption{Procedure to derive a corridor to translate enclosed nodes. We use an SVG canvas, where the point of origin (0,0) is located at the top left corner, with the x-axis extending to the right and the y-axis extending downwards (See \autoref{fig:stalemate}).}\label{alg:derive corridor}

### 2022-07-16 paper/section/related.tex
draft: appendix on QGIS pre-processing

\citea{warf2008Geography} use a Dorling cartogram to represent the religious diversity in the United States. \citea{sun2010Effectiveness} depict 1996 US election data and 2005 Chinese population data using Dorling, Mosaic, and contiguous cartograms. \citea{gao2020Visualising} present a Dorling cartogram to illustrate COVID-19 infections in China. \citea{tong2018cartograms} use a Demers cartogram to visualize health-related data by CCG regions in England, the work introduces a novel technique to remove overlap of squares based on topological features, aiming to improve both geographical and topological accuracy. \citea{nusrat2020Recognition} investigate the memorability of contiguous and Dorling cartograms by using multiple datasets including demographics, agriculture, and retail data in the US. See \tableref{table:region vs node} for a list of literature that adopts cartograms for visualization with corresponding geographical regions and node counts.

>>>>

\citea{warf2008Geography} use a Dorling cartogram to represent the religious diversity in the United States. \citea{sun2010Effectiveness} depict 1996 US election data and 2005 Chinese population data using Dorling, Mosaic, and contiguous cartograms. \citea{gao2020Visualising} present a Dorling cartogram to illustrate COVID-19 infections in China. \citea{tong2018cartograms} use a Demers cartogram to visualize health-related data by CCG regions in England, the work introduces a novel technique to remove overlap of squares based on topological features, aiming to improve both geographical and topological accuracy. \citea{nusrat2020Recognition} investigate the memorability of contiguous and Dorling cartograms by using multiple datasets including demographics, agriculture, and retail data in the US. See \autoref{table:region vs node} for a list of literature that adopts cartograms for visualization with corresponding geographical regions and node counts.
